CrypTool Script\footnote{ Background reading for the CrypTool program}\\ Mathematics and Cryptography

### 2003-07-08 script/de/chapters/numbertheory.tex
*** empty log message ***
removed nbsps

\begin{quote}
$$ (4^3)^2 \equiv 64^2 \equiv 4096 \equiv 1 {\rm ~(mod~} 5). $$
(1) Reduziert man bereits {\bf Zwischenergebnisse} modulo $5$, kann man
schneller\footnote{%
Die Rechenzeit der Multiplikation zweier Zahlen h"angt normalerweise von der
L"ange der Zahlen ab. Dies sieht man, wenn man nach der Schulmethode z.B.
$474*228$ berechnet: Der Aufwand steigt quadratisch, da $3*3$ Ziffern
multipliziert werden m"ussen. Durch die Reduktion der Zwischenergebnisse
werden die Zahlen deutlich kleiner.
} rechnen, muss aber auch aufpassen, da sich dann nicht immer alles wie in der
gew"ohnlichen Arithmetik verh"alt.
\begin{eqnarray*}
(4^3)^2 & \equiv & (4^3{\rm ~(mod~}5))^2{\rm ~(mod~}5) \nonumber \\
            & \equiv & (64{\rm ~(mod~}5))^2{\rm ~(mod~}5) \nonumber \\
            & \equiv & 4^2{\rm ~(mod~}5) \nonumber \\
            & \equiv & 16 \equiv 1 {\rm ~(mod~}5). \nonumber
\end{eqnarray*}

@@

{\bf Beispiel 2:}
$$
3^{28} \equiv 3^{4{*}7} \equiv 3^{4{*}7{\rm ~(mod~}10)} \equiv 3^8 \equiv 6561 \equiv 5 {\rm ~(mod~}11).
$$

>>>>

\begin{quote}
$$ (4^3)^2 \equiv 64^2 \equiv 4096 \equiv 1 {\rm ~(mod~} 5). $$
(1) Reduziert man bereits {\bf Zwischenergebnisse} modulo $5$, kann man
schneller\footnote{%
Die Rechenzeit der Multiplikation zweier Zahlen h"angt normalerweise von der
L"ange der Zahlen ab. Dies sieht man, wenn man nach der Schulmethode z.B.
$474*228$ berechnet: Der Aufwand steigt quadratisch, da $3*3$ Ziffern
multipliziert werden m"ussen. Durch die Reduktion der Zwischenergebnisse
werden die Zahlen deutlich kleiner.
} rechnen, muss aber auch aufpassen, da sich dann nicht immer alles wie in der
gew"ohnlichen Arithmetik verh"alt.
\begin{eqnarray*}
(4^3)^2 & \equiv & (4^3{\rm ~(mod~}5))^2{\rm ~(mod~}5) \nonumber \\
            & \equiv & (64{\rm ~(mod~}5))^2\;{\rm ~(mod~}5) \nonumber \\
            & \equiv & 4^2{\rm ~(mod~}5) \nonumber \\
            & \equiv & 16 \equiv 1 {\rm ~(mod~}5). \nonumber
\end{eqnarray*}

@@

{\bf Beispiel 2:}
$$
3^{28} \equiv 3^{4\:*\:7} \equiv 3^{4\:*\:7{\rm ~(mod~}10)} \equiv 3^8 \equiv 6561 \equiv 5 {\rm ~(mod~}11).
$$

### 2003-07-08 script/en/chapters/numbertheory.tex
*** empty log message ***
removed nbsps

{\bf Example 1:}
$$ (4^3)^2 \equiv 64^2 \equiv 4096 \equiv 1 {\rm ~(mod~} 5). $$
\begin{quote}
(1) We can speed up\footnote{The time required to calculate the multiplication 
of two numbers normally depends on the length of the numbers.  We can observe
this if we use the school method to calculate, for instance, $474*228$.  The 
time required increases 
%quadratically
in a quadratic square manner , because we need to multiply $3*3$ numbers. The 
numbers become considerably smaller if we reduce the interim result.} the 
calculation by reducing the {\bf interim results} modulo $5$ but we need 
to take care because not everything will then work in the same way as in 
standard arithmetic.
\begin{eqnarray*}
(4^3)^2 & \equiv & (4^3{\rm ~(mod~}5))^2{\rm ~(mod~}5) \nonumber \\
            & \equiv & (64{\rm ~(mod~}5))^2{\rm ~(mod~}5) \nonumber \\
            & \equiv & 4^2{\rm ~(mod~}5) \nonumber \\
            & \equiv & 16 \equiv 1 {\rm ~(mod~}5). \nonumber
\end{eqnarray*}

@@

Using $J(5) = 4$ gives:
$$
(4^3)^2 \equiv 4^{3*2{\rm ~(mod~}J(5))} \equiv 4^{6{\rm ~mod~}4} \equiv 4^2 \equiv 16 \equiv 1 {\rm 
~(mod~}5).
$$
This delivers the correct result.
\vskip + 5pt
\end{quote}

@@

{\bf Example 2:}
$$
3^{28} \equiv 3^{4*7} \equiv 3^{4*7{\rm ~(mod~}10)} \equiv 3^8 \equiv 6561 \equiv 5 {\rm ~(mod~}11).
$$

>>>>

{\bf Example 1:}
$$ (4^3)^2 \equiv 64^2 \equiv 4096 \equiv 1 {\rm ~(mod~} 5). $$
\begin{quote}
(1) We can speed up\footnote{The time required to calculate the multiplication 
of two numbers normally depends on the length of the numbers.  We can observe
this if we use the school method to calculate, for instance, $474*228$.  The 
time required increases 
%quadratically
in a quadratic square manner , because we need to multiply $3*3$ numbers. The 
numbers become considerably smaller if we reduce the interim result.} the 
calculation by reducing the {\bf interim results} modulo $5$ but we need 
to take care because not everything will then work in the same way as in 
standard arithmetic.
\begin{eqnarray*}
(4^3)^2 & \equiv & (4^3{\rm ~(mod~}5))^2{\rm ~(mod~}5) \nonumber \\
            & \equiv & (64{\rm ~(mod~}5))^2\;{\rm ~(mod~}5) \nonumber \\
            & \equiv & 4^2{\rm ~(mod~}5) \nonumber \\
            & \equiv & 16 \equiv 1 {\rm ~(mod~}5). \nonumber
\end{eqnarray*}

@@

Using $J(5) = 4$ gives:
$$
(4^3)^2 \equiv 4^{3\:*\:2{\rm ~(mod~}J(5))} \equiv 4^{6{\rm ~mod~}4} \equiv 4^2 \equiv 16 \equiv 1 {\rm 
~(mod~}5).
$$
This delivers the correct result.
\vskip + 5pt
\end{quote}

@@

{\bf Example 2:}
$$
3^{28} \equiv 3^{4\:*\:7} \equiv 3^{4\:*\:7{\rm ~(mod~}10)} \equiv 3^8 \equiv 6561 \equiv 5 {\rm ~(mod~}11).
$$

### 2003-07-08 script/en/chapters/primes.tex
removed nbsps

As early as in the 19th century, it was discovered that the $29$-digit number $$
f(7) = 2^{2^7} + 1 $$ is not prime. However, it was not until 1970 that
Morrison/Billhart managed to factorise it.
\begin{eqnarray*}\label{f7}
f(7) & = & 340,282,366,920,938,463,463,374,607,431,768,211,457 \\
& = & 59, 649, 589, 127, 497, 217 \cdot  5,704,689,200,685,129,054,721
\end{eqnarray*}

@@

% --------------------------------------------------------------------------
\hypertarget{FermatNumbers02}{}
\subsubsection[Fermat numbers $f(n) = 2^{2^n} + 1$]
              {Fermat numbers\footnotemark~$f(n) = 2^{2^n} + 1$} 
    \footnotetext{%
       The Fermat prime numbers play a role in circle division.
       As proven by Gauss\index{Gauss Carl Friedrich} a regular $p$-edge
       can only be constructed with the use of a pair of compasses and a 
       ruler, when $p$ is a Fermat prime number.
    }
    \label{L-FermatNumbers02}
    \index{Number!Fermat}\index{Fermat!number}
    \index{Prime number!Fermat} \index{Fermat!prime number}       
    As mentioned \hyperlink{FermatNumbers01}{above} in 
    chapter~\ref{FermatNumbers01}, Fermat wrote to Mersenne regarding 
    his assumption, that all numbers of this type are primes.
    Surprisingly he would have been able obtain a positive result using
    the negative prime number test for $n=5$ based on his small theorem.
$$
\begin{array}{lll}
f(0) = 2^{2^0} + 1  = 2^1 + 1 & = 3 &   \mapsto {\rm ~prime}  \\
f(1) = 2^{2^1} + 1  = 2^2 + 1 & = 5 &   \mapsto {\rm ~prime}  \\
f(2) = 2^{2^2} + 1  = 2^4 + 1 & = 17 &  \mapsto {\rm ~prime}  \\
f(3) = 2^{2^3} + 1  = 2^8 + 1 & = 257 & \mapsto {\rm ~prime}  \\
f(4) = 2^{2^4} + 1  = 2^{16} + 1 &  = \mbox{65,537} &  \mapsto {\rm ~prime}  \\
f(5) = 2^{2^5} + 1  = 2^{32} + 1 &  = \mbox{4,294,967,297} = 641 \cdot \mbox{6,700,417} &  \mapsto {\rm ~NOT~prime} ! \\
f(6) = 2^{2^6} + 1  = 2^{64} + 1 &  = \mbox{18,446,744,073,709,551,617} \\
                                 &  = \mbox{274,177} \cdot \mbox{67,280,421,310,721} & \mapsto {\rm ~NOT~prime} !\\
f(7) = 2^{2^7} + 1  = 2^{128} + 1 & = \mbox{(see page~\pageref{f7})}  
																 & \mapsto {\rm ~NOT~prime} !

>>>>

As early as in the 19th century, it was discovered that the $29$-digit number $$
f(7) = 2^{2^7} + 1 $$ is not prime. However, it was not until 1970 that
Morrison/Billhart managed to factorise it.
\begin{eqnarray*}\label{f7cole}
f(7) & = & 340,282,366,920,938,463,463,374,607,431,768,211,457 \\
& = & 59, 649, 589, 127, 497, 217 \cdot  5,704,689,200,685,129,054,721
\end{eqnarray*}

@@

% --------------------------------------------------------------------------
\hypertarget{FermatNumbers02}{}
\subsubsection[Fermat numbers $f(n) = 2^{2^n} + 1$]
              {Fermat numbers\footnotemark~$f(n) = 2^{2^n} + 1$} 
    \footnotetext{%
       The Fermat prime numbers play a role in circle division.
       As proven by Gauss\index{Gauss Carl Friedrich} a regular $p$-edge
       can only be constructed with the use of a pair of compasses and a 
       ruler, when $p$ is a Fermat prime number.
    }
    \label{L-FermatNumbers02}
    \index{Number!Fermat}\index{Fermat!number}
    \index{Prime number!Fermat} \index{Fermat!prime number}       
    As mentioned \hyperlink{FermatNumbers01}{above} in 
    chapter~\ref{FermatNumbers01}, Fermat wrote to Mersenne regarding 
    his assumption, that all numbers of this type are primes.
    Surprisingly he would have been able obtain a positive result using
    the negative prime number test for $n=5$ based on his small theorem.
$$
\begin{array}{lll}
f(0) = 2^{2^0} + 1  = 2^1 + 1 & = 3 &   \mapsto {\rm ~prime}  \\
f(1) = 2^{2^1} + 1  = 2^2 + 1 & = 5 &   \mapsto {\rm ~prime}  \\
f(2) = 2^{2^2} + 1  = 2^4 + 1 & = 17 &  \mapsto {\rm ~prime}  \\
f(3) = 2^{2^3} + 1  = 2^8 + 1 & = 257 & \mapsto {\rm ~prime}  \\
f(4) = 2^{2^4} + 1  = 2^{16} + 1 &  = \mbox{65,537} &  \mapsto {\rm ~prime}  \\
f(5) = 2^{2^5} + 1  = 2^{32} + 1 &  = \mbox{4,294,967,297} = 641 \cdot \mbox{6,700,417} &  \mapsto {\rm ~NOT~prime} ! \\
f(6) = 2^{2^6} + 1  = 2^{64} + 1 &  = \mbox{18,446,744,073,709,551,617} \\
                                 &  = \mbox{274,177} \cdot \mbox{67,280,421,310,721} & \mapsto {\rm ~NOT~prime} !\\
f(7) = 2^{2^7} + 1  = 2^{128} + 1 & = \mbox{(see page~\pageref{f7cole})}  
																 & \mapsto {\rm ~NOT~prime} !
